### 2020-02-19 main_text/main_text.tex
Main text major edits (KL) + minor edits (MM)
1st pass at incorporating changes/additions
suggested by and in response to reviewer.

\usepackage[obeyFinal]{todonotes} % todo notes and missing figures
\reversemarginpar

@@

	\end{flushleft}
	% Please keep the abstract below 300 words
	\section*{Abstract}
	Membrane transporters carry key metabolites through the cell membrane
	and, from a resource standpoint, are hypothesized to be produced when necessary. The
	expression of membrane transporters in metabolic pathways is often
	upregulated by the transporter substrate. In \emph{E. coli}, such systems
	include for example the \emph{lacY}, \emph{araFGH}, and \emph{xylFGH} genes,
	which encode for lactose, arabinose, and xylose transporters, respectively.
	As a case study of a minimal system, we build a
	generalizable physical model of the \emph{xapABR} genetic
	circuit, which features a regulatory feedback loop through membrane
	transport (positive feedback) and enzymatic degradation (negative feedback)
	of an inducer. Dynamical systems analysis and stochastic simulations show
	that the membrane transport makes the model system bistable in certain
	parameter regimes. Thus, it serves as a genetic ``on-off'' switch, enabling
	the system to only produce a set of metabolic enzymes when the corresponding
	metabolite is present in large amounts. We find that the negative feedback
	from the degradation enzyme does not significantly disturb the positive
	feedback from the membrane transporter. We investigate hysteresis in the
	switching and discuss the role of cooperativity and multiple binding sites
	in the model circuit. Fundamentally, this work explores how a stable genetic
	switch for a set of enzymes is obtained from transcriptional auto-activation
	of a membrane transporter through its substrate.

@@

	\section*{Introduction} \todo{remove images, todonotes, and funding for submission}
	Genetic regulatory circuits are fundamental building blocks of functioning
	cells and organisms. One abundant class of these circuits are genetic
	switches. Although their construction and function may differ,
	their common feature is bistability: their output gene expression
	will flow to and remain at one of two steady-state levels.
	The distribution of
	gene expression in a cell culture can then be bimodal. This is not to be
	confused with mere stochastic bimodality, where the system is not stable,
	and the gene expression in each cell can fluctuate between the two levels.

	One classic example of a genetic switch is a system where two repressor
	proteins each regulate the transcription of the
	other~\cite{Jacob1961,Gardner2000} (illustrated schematically in
	Fig~\ref{fig1:switches}). Here, one stable state is high expression of
	the first protein and low expression of the second,
	and the second stable state is the opposite. This
	switch enables the system to have a memory: if something induces expression
	of either one of the proteins, the system will remain in this state until a
	significant perturbation occurs. Another well-known and even simpler example
	is an auto-activating circuit in which a protein activates its own
	transcription~\cite{Wolf1998}. This gives the system an ``on-off'' switch.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Circuits.eps}
		\caption{{\bf A schematic of different genetic switches.}
			(A) and (B) show the two most well-known genetic switches:
			(A) two mutual repressors and (B) a self-activating gene.
			In (C), a very much simplified version of the
			circuit that we investigate in this paper can be seen, where the
			similarity to the switch in (B) is clear. A complete version of
			the model circuit can be found in Fig~\ref{fig2:model}.}
		\label{fig1:switches}
	\end{figure}

@@

	For our modeling, we focus on the \emph{xapABR} genetic circuit from
	\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
	complex. Instead of lactose, its purpose is to make use of the nucleoside
	xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
	circuit is made up of two operons: one that encodes for XapR and another
	that encodes for XapA and XapB. XapR appears to be a transcription factor
	that is induced by xanthosine and activates the \emph{xapAB} promoter,
	in close analogy to AraC, XylR, and also LacI.
	\footnote{One might object that LacI represses its target operon, while
	XapR, AraC, and XylR activate their target operons. However, the analogy
	we wish to draw is that the qualitative logic of their inducers are all
	identical, i.e., the presence of their respective inducer causes their
	target operon to be transcribed.}
	The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
	but the promoter architecture and function is not yet fully understood. The
	transcription of \emph{xapR} seems to be constitutive and not
	auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
	factors suggests that XapR appears in dimers where one dimer can bind two
	xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
	nucleoside phosphorylase that degrades xanthosine into components (ribose
	and xanthine) that can be fed into metabolic
	pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
	membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

	Experimentally it was found that the expression level of \emph{xapAB} among
	cells follows a bimodal distribution and that the system seems to be
	bistable~\cite{Chure2019}. We aim to understand which of the circuit's
	features are necessary for bistability and investigate its behavior in
	different parameter regimes. In the following section, we will discuss the
	details of our model. After estimating the free parameters we then present
	the observations we made through phase diagrams, followed by the results
	from stochastic simulations.

@@

	\begin{figure}[h!]
		\centering
		\includegraphics{media/XapSystem.eps}
		\caption{{\bf Model of the \emph{xapAB} circuit.}
			The XapR dimers are induced by xanthosine and the induced XapR binds
			cooperatively as an activator to the \emph{xapAB} promoter. For
			these two steps, quasi-equilibrium is assumed. If both XapR binding
			sites are occupied and the polymerase is bound, the gene is
			transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
			$\gamma_{\n{m}}$, and both proteins are translated at rate
			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
			Similarly, XapB is treated as a Michaelis-Menten enzyme which imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
			k_{\n{nup}}$, respectively.}
		\label{fig2:model}
	\end{figure}

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/States.eps}
		\caption{{\bf The promoter states.} We consider only the completely
			occupied state as active and all other states (faded out in the figure)
			as completely inactive. The parameters are the interaction energy of
			the two XapR dimers $\Delta E_{\n{coop}}$ and the dissociation
			constants $K_{\n{XapR}}$ and $K_{\n{P}}$ of XapR and polymerase
			to the promoter, respectively. The concentrations of polymerase
			and active XapR are denoted by $\n{[P]}$ and $\n{[XapR]}$.}
		\label{fig3:states}
	\end{figure}

	We consider only the state where both XapR binding sites are occupied as
	active and all other states as inactive, meaning they have transcription
	rate equal to zero. Experiments show that the expression becomes very weak when one of
	the XapR binding sites is removed from the promoter, suggesting that this
	simplification is reasonable~\cite{Chure2019}. Furthermore, we find that in
	the bistable parameter range, considering the single occupancy states as
	active instead has almost no influence on the results (see
	also~\nameref{S1_Text}).

@@

	\subsection*{Deterministic phase portraits}
	A standard way to analyze dynamical systems deterministically is
	to plot phase portraits. In the following, we present such plots where the
	state variables are the mRNA, the protein, and the xanthosine concentration.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Bistability.eps}
		\caption{{\bf Phase portraits showing bistability.}
			3D and 2D phase portraits for one set of parameters that leads to
			bistability. The parameter values are listed in
			Table~\ref{table1:nondim}. Note that all the concentrations
			($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
			$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The surfaces in (A) and the
			curves in (B) are the nullclines of the state variables, and their
			intersection points, marked in red in (B), are the steady-state
			solutions of the system. The region shaded in gray in (B) leads to
			negative concentrations and is unphysical. A vector plot of (B) that
			also shows the magnitude of flow at each point can be found
			in~\nameref{S1_Text}.}
		\label{fig4:bistability}
	\end{figure}

@@

	The bistability corresponds to the experimental observations\cite{Chure2019}, so the model passes this
	sanity check. Furthermore, the xanthosine and protein concentrations at the
	upper fixed point have the expected order of magnitude: the xanthosine
	concentration is roughly $10-100 \unit{mM}$, and there are roughly $500$
	proteins, which is just a bit lower than what was measured for the number of
	Nup transporters~\cite{Li2014} which fulfill a similar purpose. We do not
	have well founded expectations for the other fixed points, so no comparison
	can be made here. Nevertheless, the orders of magnitude at the lower fixed
	point -- roughly $1-10 \unit{nM}$ of xanthosine and around $5$ proteins --
	seem quite reasonable. Note that $\n{[x]_a} \approx \n{[c]_a}$ at the
	lower fixed point because there is only weak accumulation due to Nup and a
	few XapB transporters.

@@

	\paragraph*{The extracellular xanthosine concentration.}
	The parameter that is the experimentally most easily tunable and
	biologically the most relevant is the extracellular xanthosine
	concentration. When it is increased in experiments, the cells go from (1.) all
	being in the low expression state to (2.) the population being in a mixed state with some cells in a low expression state and others in a high expression state (all-or-none phenomenon) and then to (3.) all being in the high
	expression state~\cite{Chure2019}. If our model is correct, it should
	exhibit the same qualitative behavior. Indeed we find exactly this: as can
	be seen in Fig~\ref{fig5:extraxanth}, increasing $\n{[c]_a}$ makes the high
	stable fixed point appear and then, for even higher $\n{[c]_a}$, the lower
	one disappears. Thus, for low $\n{[c]_a}$ the only stable point of the system
	is at low expression, and for high $\n{[c]_a}$ there is only high
	expression. In between, there are two deterministically stable expression
	levels.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/VaryingC.eps}
		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
			All parameters but $\n{[c]_a}$ are as presented in
			Table~\ref{table1:nondim}. The extracellular xanthosine
			concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
			so $\n{[c]_a}$ is dimensionless).
			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline), but the blue curve (mRNA nullcline) is unchanged (see
			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
			only the lower fixed point (fixed point number 1), whereas in (B)
			there is only the upper one (fixed point number 3). In between lies
			the bistable case that was shown in Fig~\ref{fig4:bistability}.}
		\label{fig5:extraxanth}
	\end{figure}

@@

	\begin{figure}%[h!]
		\includegraphics{media/XapAB.eps}
		\centering
		\caption{{\bf Phase portraits without XapA or XapB.}
			All parameters are as presented in Table~\ref{table1:nondim}. In
			(A), the XapA term was removed from the kinetic equations. In (B),
			the equations lack the two terms from XapB. These plots clearly show
			that XapA has almost no influence on the qualitative behavior of the
			system (i.e. bistability and the order of magnitudes), but XapB is
			the essential feature for bistability.}
		\label{fig6:xapAB}
	\end{figure}

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/FewerSites.eps}
		\caption{{\bf Phase portraits for less or no cooperativity.}
			Most parameters are as presented in Table~\ref{table1:nondim},
			changes are mentioned below. Fixed points are marked in red. In (A),
			there is only one xanthosine binding site on XapR and everything
			unchanged for the XapR-promoter binding. Two parameters are changed:
			$\rho = 0.07$ and $\n{[c]_a} = 6$. This is necessary to compensate
			for the weaker induction such that the system is bistable. In (B),
			there is only one XapR binding site on the promoter and everything
			is unchanged for the xanthosine-XapR binding. Two parameters are
			changed: $\rho = 0.13$ and $\n{[c]_a} = 3$. In (C), there is only
			one xanthosine binding site on XapR and also only one XapR binding
			site on the the promoter. Two parameters are changed: $\rho = 0.1$
			and $\n{[XapR]_R} = 5$. Whereas bistability is retained in (A) and
			(B), it cannot be obtained anymore in (C).}
		\label{fig7:coop}
	\end{figure}

@@

	\paragraph*{Bimodality and the extracellular xanthosine concentration.}
	This stochastic approach results in the same bimodal distributions that were already
	seen in the deterministic investigation and experimental studies.
	Fig~\ref{fig8:stochC} shows the distribution of protein expression found in
	the simulations for different values of the extracellular xanthosine
	concentration. The parameters that were used are the same as in the previous
	section (listed in Table~\ref{table1:nondim}). To obtain the distributions,
	we ran the simulation 5000 times for a simulated time of $10^6 \unit{s}$
	each and started at a mRNA, protein and intracellular xanthosine count of 0.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/HistogramsC.eps}
		\caption{{\bf Distributions from stochastic simulations and the corresponding phase portraits.}
			Apart from $\n{[c]_a}$, the parameters are the same as in
			Table~\ref{table1:nondim}. For the distributions, the simulations
			were run 5000 times for $10^6 \unit{s}$ each (simulated time) and
			started at a mRNA, protein and intracellular xanthosine count of 0.
			We show the two cases of unimodality (low expression in (A) and high
			expression in (C)) as well as the case of bimodality in (B). The
			values of $\n{[c]_a}$ are 12 in (A), 18.5 in (B), and 25 in (C)
			(recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
			so $\n{[c]_a}$ is dimensionless). The output from the stochastic simulations
			is in good agreement with the concentrations at the fixed points in
			the deterministic phase portraits.}
		\label{fig8:stochC}
	\end{figure}

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations.

@@

	\paragraph*{Time evolution and switching times.}
	Fig~\ref{fig:StochT} shows the time evolution of the protein concentration for one exemplary run of the simulation.
	Again, the simulation was started with a mRNA, protein and intracellular
	xanthosine count of 0 and was run for a simulated time of $5 \cdot 10^5
	\unit{s}$. In this specific example, switching occurred after $10^5
	\unit{s}$.

	Comparing this to the experimentally expected timescales~\cite{Chure2019} is difficult,
	because the switching time strongly depends on the extracellular xanthosine
	concentration. Experiments were always stopped after a few hours, and in
	this time, the cell population might not reach its steady-state expression
	distribution. Hence, the distribution could be bimodal when the experiment
	is stopped but become unimodal after further waiting. That way,
	extracellular xanthosine concentrations that are too high for deterministic
	bistability could lead to experimental bimodality if the experiment is
	stopped too early. In this case, the observed switching time would be
	shorter, which makes the comparison to our simulations even harder. Thus, we
	cannot say if it is problematic that the $10^5 \unit{s}$ is larger than what
	was found in the experiment.

>>>>

\reversemarginpar

@@

	\end{flushleft}
	% Please keep the abstract below 300 words
	\section*{Abstract}
	Membrane transporters carry key metabolites through the cell membrane
	and, from a resource standpoint, are hypothesized to be produced only when necessary. The
	expression of membrane transporters in metabolic pathways is often
	upregulated by the transporter substrate. In \emph{E. coli}, such systems
	include for example the \emph{lacY}, \emph{araFGH}, and \emph{xylFGH} genes,
	which encode for lactose, arabinose, and xylose transporters, respectively.
	As a case study of a minimal system, we build a
	generalizable physical model of the \emph{xapABR} genetic
	circuit, which features a regulatory feedback loop through membrane
	transport (positive feedback) and enzymatic degradation (negative feedback)
	of an inducer. Dynamical systems analysis and stochastic simulations show
	that the membrane transport makes the model system bistable in certain
	parameter regimes. Thus, it serves as a genetic ``on-off'' switch, enabling
	the cell to only produce a set of metabolic enzymes when the corresponding
	metabolite is present in large amounts. We find that the negative feedback
	from the degradation enzyme does not significantly disturb the positive
	feedback from the membrane transporter. We investigate hysteresis in the
	switching and discuss the role of cooperativity and multiple binding sites
	in the model circuit. Fundamentally, this work explores how a stable genetic
	switch for a set of enzymes is obtained from transcriptional auto-activation
	of a membrane transporter through its substrate.

@@

	\section*{Introduction} 
	Genetic regulatory circuits are fundamental building blocks of functioning
	cells and organisms. One abundant class of these circuits are genetic
	switches. Although their construction and function may differ,
	their common feature is bistability: their output gene expression
	will flow to and remain at one of two steady-state levels.
	The distribution of
	gene expression in a cell culture can then be bimodal. This is not to be
	confused with mere stochastic bimodality, where the system is not stable,
	and the gene expression in each cell can fluctuate between the two levels.

	One classic example of a genetic switch is a system where two repressor
	proteins each regulate the transcription of the
	other~\cite{Jacob1961,Gardner2000} (illustrated schematically in
	Fig~\ref{fig0:switches}). Here, one stable state is high expression of
	the first protein and low expression of the second,
	and the second stable state is the opposite. This
	switch enables the system to have a memory: if something induces expression
	of either one of the proteins, the system will remain in this state until a
	significant perturbation occurs. Another well-known and even simpler example
	is an auto-activating circuit in which a protein activates its own
	transcription~\cite{Wolf1998}. This gives the system an ``on-off'' switch.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig0_switches.eps}
		\caption{{\bf A schematic of different genetic switches.}
			(A) and (B) show the two most well-known genetic switches:
			(A) two mutual repressors and (B) a self-activating gene.
			In (C), a very much simplified version of the
			circuit that we investigate in this paper can be seen, where the
			similarity to the switch in (B) is clear. A complete version of
			the model circuit can be found in Fig~\ref{fig2:model}.}
		\label{fig0:switches}
	\end{figure}

@@

	For our modeling, we focus on the \emph{xapABR} genetic circuit from
	\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
	complex. Instead of lactose, its purpose is to make use of the nucleoside
	xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
	circuit is made up of two operons: one that encodes for XapR and another
	that encodes for XapA and XapB. XapR appears to be a transcription factor
	that is induced by xanthosine and activates the \emph{xapAB} promoter,
	in close analogy to AraC, XylR, and also LacI.\footnote{One might object that LacI represses its target operon, while
	XapR, AraC, and XylR activate their target operons. However, the analogy
	we wish to draw is that the qualitative logic of their inducers are all
	identical, i.e., the presence of their respective inducer causes their
	target operon to be transcribed.} The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
	but the promoter architecture and function is not yet fully understood. The
	transcription of \emph{xapR} seems to be constitutive and not
	auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
	factors suggests that XapR appears in dimers where one dimer can bind two
	xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
	nucleoside phosphorylase that degrades xanthosine into components (ribose
	and xanthine) that can be fed into metabolic
	pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
	membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

	Experimentally, we found that the expression level of \emph{xapAB} among
	cells follows a bimodal distribution and that the system seems to be
	bistable (see next section). We aim to understand which of the circuit's
	features are necessary for bistability and investigate its behavior in
	different parameter regimes. After presenting some experimental background on the \emph{xap} genes, we will discuss the
	details of our model. Next, we estimate the free parameters and then present
	the observations we made through phase diagrams, followed by the results
	from stochastic simulations.

	\section*{Experimental motivation} \label{sec:experiment}
	Our work was motivated by the experimental observation of bimodality in the \emph{xap} circuit, which is shown in Fig~\ref{fig1:data}. We focus on the essential findings here and refer the reader to~\nameref{S1_Text} for more experimental details. Briefly, we placed a fluorescent reporter under the control of the wild type (wt) \emph{xapAB} promoter. This construct was placed in three different backgrounds: \textDelta\emph{xapABR}, \textDelta\emph{xapAB}, and wt. Fig~\subref{fig1:data}{A} shows that for increasing xanthosine concentration, expression in the wt-background increases in a switch-like way: it is nearly zero when there is little xanthosine (``uninduced state''), but increases drastically when there is more (``induced state''). In between, the aforementioned bimodal distribution is obtained, where some bacteria are in the uninduced expression state and others are induced. \textcolor{red}{maybe remove: In addition, we observed that the property of being induced seems to be passed on to daughter cells, suggesting that there is not merely bimodality, but bistability.}

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.4\textwidth]{media/Fig1_data.pdf}
		\caption{{\bf Experimental data on the \emph{xap} circuit.} The expression of the \emph{xapAB} promoter was measured for different concentrations of xanthosine in the bacteria's environment. (A) shows the wild type circuit while in (B) the genes \emph{xapABR} and in (C) the genes \emph{xapAB} are removed. The wild type circuit behaves like a switch.}
		\label{fig1:data}
	\end{figure}

	When all genes of the \emph{xapABR} circuit are removed, the xanthosine response of our reporter construct disappears (see Fig~\subref{fig1:data}{B}, where the last line is autofluorescence). This is the result of removing the transcription factor XapR that is induced by xanthosine. When only the genes \emph{xapAB} are removed but \emph{xapR} is kept, a response to the xanthosine concentration is regained but it is no longer switch-like (see Fig~\subref{fig1:data}{C}). Instead, the distribution remains unimodal and simply shifts to higher expression as the xanthosine concentration is increased. As we will see later on, this is because the circuit now lacks the positive feedback loop due to the xanthosine membrane transporter XapB.

	Clearly, XapR regulates the \emph{xapAB} promoter. As mentioned in the introduction, there are two binding sites for XapR on the \emph{xapAB} promoter: one directly at the polymerase binding site, and one further upstream. When the upstream binding site is removed, bimodality is lost and the expression of \emph{xapAB} becomes several orders of magnitude weaker, as can be seen in Fig~\ref{}.
	\textcolor{red}{data missing}

	Already, we believe this data, combined with the prior knowledge of the field, clearly suggest a minimal model of the \emph{xapABR} system. The construction and exploration of this model occupies the remainder of this work. We leave it for future work to quantitatively dissect and test the model, in the manner of, for example, \textcolor{red}{citations}.

@@

	\begin{figure}[h!]
		\centering
		\includegraphics{media/Fig2_model.eps}
		\caption{{\bf Model of the \emph{xapAB} circuit.}
			The XapR dimers are induced by xanthosine and the induced XapR binds
			cooperatively as an activator to the \emph{xapAB} promoter. For
			these two steps, quasi-equilibrium is assumed. If both XapR binding
			sites are occupied and the polymerase is bound, the gene is
			transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
			$\gamma_{\n{m}}$, and both proteins are translated at rate
			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
			Similarly, XapB is treated as a Michaelis-Menten enzyme which imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
			k_{\n{nup}}$, respectively.}
		\label{fig2:model}
	\end{figure}

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig3_states.eps}
		\caption{{\bf The promoter states.} We consider only the completely
			occupied state as active and all other states (faded out in the figure)
			as completely inactive. The parameters are the interaction energy of
			the two XapR dimers $\Delta E_{\n{coop}}$ and the dissociation
			constants $K_{\n{XapR}}$ and $K_{\n{P}}$ of XapR and polymerase
			to the promoter, respectively. The concentrations of polymerase
			and active XapR are denoted by $\n{[P]}$ and $\n{[XapR]}$.}
		\label{fig3:states}
	\end{figure}

	We consider only the state where both XapR binding sites are occupied as
	active and all other states as inactive, meaning they have transcription
	rate equal to zero. Our experiments show that the expression becomes very weak when one of
	the XapR binding sites is removed from the promoter (see~\nameref{S1_Text}), suggesting that this
	simplification is reasonable. Furthermore, we find that in
	the bistable parameter range, considering the single occupancy states as
	active instead has almost no influence on the results (see
	also~\nameref{S1_Text}).

@@

	\subsection*{Deterministic phase portraits}
	A standard way to analyze dynamical systems deterministically is
	to plot phase portraits. In the following, we present such plots where the
	state variables are the mRNA, the protein, and the xanthosine concentration. Note that for the low copy numbers that can occur in our system, a deterministic analysis is not necessarily meaningful. Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results. Thus, deterministic phase portraits are a good starting point.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig4_bistability.eps}
		\caption{{\bf Phase portraits showing bistability.}
			3D and 2D phase portraits for one set of parameters that leads to
			bistability. The parameter values are listed in
			Table~\ref{table1:nondim}. Note that all the concentrations
			($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
			$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The surfaces in (A) and the
			curves in (B) are the nullclines of the state variables, and their
			intersection points, marked in red in (B), are the steady-state
			solutions of the system. The region shaded in gray in (B) leads to
			negative concentrations and is unphysical. A vector plot of (B) that
			also shows the magnitude of flow at each point can be found
			in~\nameref{S1_Text}.}
		\label{fig4:bistability}
	\end{figure}

@@

	The bistability corresponds to the experimental observations, so the model passes this
	sanity check. Furthermore, the xanthosine and protein concentrations at the
	upper fixed point have the expected order of magnitude: the xanthosine
	concentration is roughly $10-100 \unit{mM}$, and there are roughly $500$
	proteins, which is just a bit lower than what was measured for the number of
	Nup transporters~\cite{Li2014} which fulfill a similar purpose. We do not
	have well founded expectations for the other fixed points, so no comparison
	can be made here. Nevertheless, the orders of magnitude at the lower fixed
	point -- roughly $1-10 \unit{nM}$ of xanthosine and around $5$ proteins --
	seem quite reasonable. Note that $\n{[x]_a} \approx \n{[c]_a}$ at the
	lower fixed point because there is only weak accumulation due to Nup and a
	few XapB transporters.

@@

	\paragraph*{The extracellular xanthosine concentration.}
	The parameter that is the experimentally most easily tunable and
	biologically the most relevant is the extracellular xanthosine
	concentration. When it is increased in experiments, the cells go from (1.) all
	being in the low expression state to (2.) the population being in a mixed state with some cells in a low expression state and others in a high expression state (all-or-none phenomenon) and then to (3.) all being in the high
	expression state (see section ``experimental motivation''). If our model is correct, it should
	exhibit the same qualitative behavior. Indeed we find exactly this: as can
	be seen in Fig~\ref{fig5:extraxanth}, increasing $\n{[c]_a}$ makes the high
	stable fixed point appear and then, for even higher $\n{[c]_a}$, the lower
	one disappears. Thus, for low $\n{[c]_a}$ the only stable point of the system
	is at low expression, and for high $\n{[c]_a}$ there is only high
	expression. In between, there are two deterministically stable expression
	levels. Later on in this text, we show this result in form of a bifurcation diagram.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig5_extraxanth.eps}
		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
			All parameters but $\n{[c]_a}$ are as presented in
			Table~\ref{table1:nondim}. The extracellular xanthosine
			concentration in these plots is $\n{[c]_a}=7$ in (A) and
			$\n{[c]_a}= 40$ in (B) (recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
			so $\n{[c]_a}$ is dimensionless).
			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline), but the blue curve (mRNA nullcline) is unchanged (see
			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
			only the lower fixed point (fixed point number 1), whereas in (B)
			there is only the upper one (fixed point number 3). In between lies
			the bistable case that was shown in Fig~\ref{fig4:bistability}.}
		\label{fig5:extraxanth}
	\end{figure}

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig6_xapAB.eps}
		\caption{{\bf Phase portraits without XapA or XapB.}
			All parameters are as presented in Table~\ref{table1:nondim}. In
			(A), the XapA term was removed from the kinetic equations. In (B),
			the equations lack the two terms from XapB. These plots clearly show
			that XapA has almost no influence on the qualitative behavior of the
			system (i.e. bistability and the order of magnitudes), but XapB is
			the essential feature for bistability.}
		\label{fig6:xapAB}
	\end{figure}

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig7_coop.eps}
		\caption{{\bf Phase portraits for less or no cooperativity.}
			Most parameters are as presented in Table~\ref{table1:nondim},
			changes are mentioned below. Fixed points are marked in red. In (A),
			there is only one xanthosine binding site on XapR and everything
			unchanged for the XapR-promoter binding. Two parameters are changed:
			$\rho = 0.07$ and $\n{[c]_a} = 6$. This is necessary to compensate
			for the weaker induction such that the system is bistable. In (B),
			there is only one XapR binding site on the promoter and everything
			is unchanged for the xanthosine-XapR binding. Two parameters are
			changed: $\rho = 0.13$ and $\n{[c]_a} = 3$. In (C), there is only
			one xanthosine binding site on XapR and also only one XapR binding
			site on the the promoter. Two parameters are changed: $\rho = 0.1$
			and $\n{[XapR]_R} = 5$. Whereas bistability is retained in (A) and
			(B), it cannot be obtained anymore in (C).}
		\label{fig7:coop}
	\end{figure}

@@

	Note that we neglect stochastic fluctuations in any of our parameters, in particular in the XapR copy number. The latter is on the order of 10, but because of the long life span and rare expression of proteins like XapR, we expect the influence of the simplification on our results to be rather small. Nevertheless, the overall stochastic fluctuations are expected to be larger in the real system than in our simulations.

	\paragraph*{Bimodality and the extracellular xanthosine concentration.}
	The stochastic approach results in the same bimodal distributions that were already
	seen in the deterministic investigation and experimental studies.
	Fig~\ref{fig8:stochC} shows the distribution of protein expression found in
	the simulations for different values of the extracellular xanthosine
	concentration. The parameters that were used are the same as in the previous
	section (listed in Table~\ref{table1:nondim}). However, we also find good agreement of the simulations and the deterministic results for other sets of parameters. To obtain the distributions,
	we ran the simulation 5000 times for a simulated time of $10^6 \unit{s}$
	each and started at a mRNA, protein and intracellular xanthosine count of 0.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig8_distribution.eps}
		\caption{{\bf Distributions from stochastic simulations and the corresponding phase portraits.}
			Apart from $\n{[c]_a}$, the parameters are the same as in
			Table~\ref{table1:nondim}. For the distributions, the simulations
			were run 5000 times for $10^6 \unit{s}$ each (simulated time) and
			started at a mRNA, protein and intracellular xanthosine count of 0.
			We show the two cases of unimodality (low expression in (A) and high
			expression in (C)) as well as the case of bimodality in (B). The
			values of $\n{[c]_a}$ are 12 in (A), 18.5 in (B), and 25 in (C)
			(recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
			so $\n{[c]_a}$ is dimensionless). The output from the stochastic simulations
			is in good agreement with the concentrations at the fixed points in
			the deterministic phase portraits.}
		\label{fig8:stochC}
	\end{figure}

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. Also note the similarity to the shape of the experimental distributions that were presented in section ``experimental motivation''. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations. This actually is a good finding: the biologically interesting feature is the switch-like behavior, whereas bimodality can be problematic for biology as this means that there is a large insecurity in the input-output behavior of a bacteria's reaction to xanthosine. Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch.

@@

	\paragraph*{Time evolution and adaptation times.}
	Fig~\ref{fig9:StochT} shows the time evolution of the protein concentration for one exemplary run of the simulation.
	Again, the simulation was started with a mRNA, protein and intracellular
	xanthosine count of 0 and was run for a simulated time of $5 \cdot 10^5
	\unit{s}$. In this specific example, the fixed point was reached after $10^5
	\unit{s}$. Of course, the system moves to its mean steady-state much faster when the extracellular xanthosine concentration is further away from the bimodal regime (in analogy to a ``critical point''). For comparison, the trajectory obtained from integrating the deterministic system is also shown in Fig~\ref{fig9:StochT}. It clearly agrees well with the simulation. There are some cases where the big adaptation step happens slightly earlier or later, but still at around $10^5 \unit{s}$.
